Modal Abstractions for Virtualizing Memory Addresses

### 2024-07-10 examples.tex
+

@@

Concretely speaking, when we focus on the specfication of Line 13 in Figure \ref{fig:calltopteinitialize} in Figure \ref{fig:lookingupidpampping}, to read the value referenced by physical address \textsf{entry}, as shown in the first specification Line in Figure \ref{fig:lookingupidpampping}, we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the fractional ownership of the ghost translation and physical pointsto
% \[\sumwalkabsent \ast \mathsf{entry}\mapsto_{\mathsf{p,qfrac}}\; \vale \]
\[\fracghostmaptoken{\delta}{(\texttt{entry + KERNBASE})}{\textsf{entry}}{\textsf{qfrac}} \ast \mathsf{entry}\mapsto_{\mathsf{p,qfrac}}\; \vale \]
by using the capability $\Xi \;!!\; \textsf{entry} = \textsf{Some V}$ and left the invariant as in the state $\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi\setminus\{\mathsf{entry}\},m)$ , and construct the virtual mapping with the resources peeled-out of the identity mapping $\Xi$ 
\[ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \]
under the side condition (*)
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
assuring that looking the identity mapping for \textsf{entry} is safe under the subtle justification which equates the full ownership to the non/presence of the entry which can only be known when investigated ad Line 19in Figure \ref{fig:calltopteinitialize}.
 \begin{figure*}
   \footnotesize
   \[
 \begin{array}{l}
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast \ulcorner \Xi !! \textsf{entry} = \textsf{Some V} \urcorner  }_{\rtv}  \\
\sqsubseteq \\
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
\mathsf{mov} \;   [ \mathsf{r8} ],\mathsf{rdi}\\
 \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
 \end{array}
 \]
 \vspace{-1em}
\caption{Peel-out the resources needed for the virtual-pte pointsto relation for physical address \textsf{entry} to be accessed in Line 7 in Figure \ref{fig:calltopteinitialize}. }
  \label{fig:lookingupidpampping}
 \end{figure*}

 \subsubsection{Ensuring Presence of an Entry}
 After obtaining the identity mapping for \textsf{entry}, we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit through
 \[ \begin{array}{l}
   \mathsf{and }\; 0x1, \;rdi \\
   \mathsf{mov } \;\mathsf{rdi}, \; \mathsf{rax} \\
   \mathsf{cmp } \; 0x0, \;\mathsf{rax}
 \end{array}
 \] Lines (17-20) in Figure \ref{fig:calltopteinitialize}.
 Accessing the presence bit and checking the value allow  us to exploit the fact in the side condition (*) in such a way that we either take the allocation path in the non-presence case (Lines 19-44 in Figure \ref{fig:calltopteinitialize}) with the knowledge of full ownership ($\textsf{qfrac} = 1$) to set the entry with proper values for the bitfields, most imporantly frame referencing freshly allocated page and presence bit, or, take non-allocation path and jump to Line 45 with fractional permission for looking up the entry values.

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{\_} \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_}\ast\ghostmaptoken{\delta{}s}{\rtv}{\delta} }_{\rtv}$
;;void pte_initialize(pte_t **table) {
...
;;set up the stack    
;;allocate a full zeroed page for 512 8-byte entries 
;;pte_t *local = kalloc();
callq  81 <pte_initialize+0x11>	7d: R_X86_64_PLT32	kalloc-0x4
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_} \ast \textsf{nextpaddr} \mapsto_{\textsf{p}} \textsf{w0\_64}\ast\ghostmaptoken{\delta{}s}{\rtv}{\delta} }_{\rtv}$
mov    rax,-0x10[rbp] 
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \ulcorner \Xi \;!!\; \mathsf{nextpaddr} = Z \urcorner \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_} \ast \textsf{nextpaddr} \mapsto_{\textsf{p}} \textsf{w0\_64} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta} }_{\rtv}$  
;;*table = local;
mov    -0x10[rbp],rax
mov    -0x8[rbp],rdi
mov    rax,[rdi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr}  \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\mathsf{nextpaddr}} \ast \textsf{nextpaddr} \mapsto_{\textsf{p}} \textsf{w0\_64} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta} }_{\rtv}$  
;;return;
;;..
;;clean up the stack
\end{lstlisting}
\vspace{-1em}
\caption{Allocating a physical page }
\label{pteinitializespec}
\end{figure}
\todo{ISMAIL: Change the following paragraphs and LInes 33 35  in \ref{fig:calltopteinitialize}, because wpdp2V is where we only know the relation between entry->pfn nextpaddr , consequently zeroed. Not here }
Since we are using \textsf{pte\_initialize} for page-table address allocation, we must relate this newly allocated physical address to the identity mapping map $\Xi$ -- Specification Lines 33 and 35 in Figure \ref{fig:calltopteinitialize}. There are two important aspects on relating a freshly allocated page-table physical address to the identity-mapping which we focus in more detail separately in the specification of \textsf{pte\_initialize} in Figure \ref{fig:pteinitializespec}. First, one might have expected to see a physical pointsto for the fresly page-table address  ($\mathsf{nextpaddr} \mapsto_{\mathsf{p}} \mathsf{w64\_0}$) after the call to \textsf{pte\_initialize} but we do not have this physical pointsto resource. Second, we see the identity mapping for \textsf{nextpaddr}, first not appearing in the identity map (Specification Line 11 in Figure \ref{fig:pteinitializespec}), then as inserted into the identity-map with a value \textsf{Z} ($\Xi \cup \{\mathsf{nextpaddr}\}$) together with the fragmental capability  -- $\ulcorner \Xi \; !!\; \textsf{nextpaddr} = \textsf{ Some Z} \urcorner$ -- at Lines 13 and 14 in Figure \ref{fig:pteinitializespec} . The value type \textsf{Z}, unlike the value type \textsf{V} denoting the random value in Line 8 in Figure \ref{fig:calltopteinitialize}, denotes \textsf{zeroed} value.

>>>>

  \begin{lstlisting}[mathescape, basicstyle=\tiny]

@@

Concretely speaking, when we focus on the specfication of Line 13 in Figure \ref{fig:calltopteinitialize} in Figure \ref{fig:lookingupidpampping}, to read the value referenced by physical address \textsf{entry}, as shown in the first specification Line in Figure \ref{fig:lookingupidpampping}, we open our the addresss-space invariant ($\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi,m)$), and then obtain the fractional ownership of the ghost translation and physical pointsto
% \[\sumwalkabsent \ast \mathsf{entry}\mapsto_{\mathsf{p,qfrac}}\; \vale \]
\[\fracghostmaptoken{\delta}{(\texttt{entry + KERNBASE})}{\textsf{entry}}{\textsf{qfrac}} \ast \mathsf{entry}\mapsto_{\mathsf{p,qfrac}}\; \vale \]
by using the capability $\Xi \;!!\; \textsf{entry} = \textsf{Some V}$ and left the invariant as in the state $\mathcal{I}\textsf{ASpace}_{\textsf{id}}(\ptablestore,\Xi\setminus\{\mathsf{entry}\},m)$ , and construct the virtual mapping with the resources peeled-out of the identity mapping $\Xi$ 
\[ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \]
under the side condition (*)
\begin{equation*}
 \ulcorner \textsf{qfrac} = 1 \leftrightarrow \; \lnot\textsf{entry\_present }(\vale) \urcorner \tag{*}
\end{equation*}
assuring that looking the identity mapping for \textsf{entry} is safe under the subtle justification which equates the full ownership to the non/presence of the entry which can only be known when investigated in Line 19in Figure \ref{fig:calltopteinitialize}.
 \begin{figure*}
   \footnotesize
 \begin{array}{l}
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m) \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry} \textsf{\_} \ast \ulcorner \Xi !! \textsf{entry} = \textsf{Some V} \urcorner  }_{\rtv}  \\
\sqsubseteq \\
   \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
\mathsf{mov} \;   [ \mathsf{r8} ],\mathsf{rdi}\\
 \specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi\setminus \{ \mathsf{entry} \},m)  \ast \textsf{rbp-8} \mapsto_{\textsf{v}} \textsf{entry} \ast \texttt{r8}  \mapsto_{\textsf{r}} \textsf{entry}  \ast  \texttt{rbp-16} \mapsto_{\textsf{v}} \textsf{next} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{entry} \ast \ghostmaptoken{\delta{}s}{\rtv}{\delta}  }_{\rtv} \\
\specline{ \textsf{entry+KERNBASE} \mapsto_{\textsf{vpte,qfrac}} \textsf{entry entry\_val} \ast \ulcorner  \textsf{qfrac} = 1 \leftrightarrow \; \lnot(\textsf{entry\_present entry\_val})\urcorner}_{\rtv} \\
 \end{array}
 \]
\caption{Peel-out the resources needed for the virtual-pte pointsto relation for physical address \textsf{entry} to be accessed in Line 7 in Figure \ref{fig:calltopteinitialize}. }
  \label{fig:lookingupidpampping}
 \end{figure*}
 \subsubsection{Ensuring Presence of an Entry}
 After obtaining the identity mapping for \textsf{entry}, we are able to load the \textsf{entry\_val} into \textsf{rdi}, and check the presence bit through
 \[ \begin{array}{l}
   \mathsf{and }\; 0x1, \;rdi \\
   \mathsf{mov } \;\mathsf{rdi}, \; \mathsf{rax} \\
   \mathsf{cmp } \; 0x0, \;\mathsf{rax}
 \end{array}
 \] Lines (17-20) in Figure \ref{fig:calltopteinitialize}.
 Accessing the presence bit and checking the value allow  us to exploit the fact in the side condition (*) in such a way that we either take the allocation path in the non-presence case (Lines 19-44 in Figure \ref{fig:calltopteinitialize}) with the knowledge of full ownership ($\textsf{qfrac} = 1$) to set the entry with proper values for the bitfields, most imporantly frame referencing freshly allocated page and presence bit, or, take non-allocation path and jump to Line 45 with fractional permission for looking up the entry values.

@@

$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{r}} \textsf{\_} \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\_}\ast\ghostmaptoken{\delta{}s}{\rtv}{\delta} }_{\rtv}$
;;void pte_initialize(pte_t **table) {
...
;;set up the stack    
;;allocate a full zeroed page for 512 8-byte entries 
;;pte_t *local = kalloc();
callq  81 <pte_initialize+0x11>	7d: R_X86_64_PLT32	kalloc-0x4
mov    rax,-0x10[rbp] 
;;*table = local;
mov    -0x10[rbp],rax
mov    -0x8[rbp],rdi
mov    rax,[rdi]
$\specline{\textsf{P} \ast \mathcal{I}\texttt{ASpace}_{\textsf{id}}(\theta,\Xi,m)   \ast \texttt{rbp-16} \mapsto_{\textsf{r}} \textsf{\_} \ast \texttt{rdi}  \mapsto_{\textsf{r}} \textsf{table\_addr}  }_{\rtv}$
$\specline{\texttt{rax} \mapsto_{\textsf{v}} \textsf{nextpaddr} \ast \textsf{table\_addr} \mapsto_{\mathsf{v}} \mathsf{\mathsf{nextpaddr}} \ast \textsf{nextpaddr} \mapsto_{\textsf{id}} \textsf{\_}\ast \forall_{i\in\textsf{0..511}} \textsf{nextpaddr + i * 8} \mapsto_{\textsf{id}} \textsf{\_} \ast\ghostmaptoken{\delta{}s}{\rtv}{\delta} }_{\rtv}$
;;return;
;;..
;;clean up the stack
\end{lstlisting}
\vspace{-1em}
\caption{Allocating a physical page }
\label{pteinitializespec}
\end{figure}
\todo{ISMAIL: Change the following paragraphs and LInes 33 35  in \ref{fig:calltopteinitialize}, because wpdp2V is where we only know the relation between entry->pfn nextpaddr , consequently zeroed. Not here }
Since we are using \textsf{pte\_initialize} for page-table address allocation, we must relate this newly allocated physical address to the identity mapping map $\Xi$ -- Specification Lines 33 and 35 in Figure \ref{fig:calltopteinitialize}. There are two important aspects on relating a freshly allocated page-table physical address to the identity-mapping which we focus in more detail separately in the specification of \textsf{pte\_initialize} in Figure \ref{fig:pteinitializespec}. First, one might have expected to see a physical pointsto for the fresly page-table address  ($\mathsf{nextpaddr} \mapsto_{\mathsf{p}} \mathsf{w64\_0}$) after the call to \textsf{pte\_initialize} but we do not have this physical pointsto resource. Second, we see the identity mapping for \textsf{nextpaddr}, first not appearing in the identity map (Specification Line 11 in Figure \ref{fig:pteinitializespec}), then as inserted into the identity-map with a value \textsf{Z} ($\Xi \cup \{\mathsf{nextpaddr}\}$) together with the fragmental capability  -- $\ulcorner \Xi \; !!\; \textsf{nextpaddr} = \textsf{ Some Z} \urcorner$ -- at Lines 13 and 14 in Figure \ref{fig:pteinitializespec} . The value type \textsf{Z}, unlike the value type \textsf{V} denoting the random value in Line 8 in Figure \ref{fig:calltopteinitialize}, denotes \textsf{zeroed} value.
